---
title: "1. Download and process NLDAS"
author: "Phil Savoy"
date: "3/12/2026"
output: html_document
---

# Introduction

This article describes downloading and processing total incoming shortwave radiation (W m^-2^) using the **StreamLightUtils** package. Total incoming shortwave radiation (W m^-2^) is provided by the National Land Data Assimilation System (NLDAS) at hourly timesteps at 0.125 degree spatial resolution. There are two potential workflows available: 1.) download data for a single site, and 2.) bulk download of data for many sites.

## Earthdata login {-}

Previously, this packaged used the NASA data rods service to download NLDAS data. However, this service has been retired and a new workflow is required.

**Downloading NLDAS data now requires an Earthdata login.** If you do not already have an account, you can [register for a new Earthdata account](https://urs.earthdata.nasa.gov/users/new). Once you have an account, you may [login to your account](https://urs.earthdata.nasa.gov/). From your account page, click on the "generate token" tab and paste the token string into a txt file and save it somewhere locally as token.txt. **Note** please be careful where you save this token and make sure not to accidentally include it in projects that are managed on places like GitHub. Unfortunately, you will have to generate a new token every 60 days as they expire.

You can then read the access token into R, modifying <path_to_token> to reflect the path where your token is saved.

```{r echo=TRUE, results='hide', eval = FALSE}
#Read in Earthdata access token
  access_token <- readLines("<path_to_token>/token.txt")
```


## 1. Downloading and processing NLDAS data for a single site

Incoming shortwave radiation data at a single site can be downloaded using the function **<span style="color:DarkOrange">nldas_get</span>** which has the following structure:

**<span style="color:DarkOrange">nldas_get</span>**(*<span style="color:#009faf">Site_ID</span>*, *<span style="color:#009faf">Lat</span>*, *<span style="color:#009faf">Lon</span>*, *<span style="color:#009faf">startDate</span>*, *<span style="color:#009faf">endDate</span>*, *<span style="color:#009faf">access_token</span>*, *<span style="color:#009faf">save_dir</span>*, )

- *<span style="color:#009faf">Site_ID</span>*  The site identifier ("Site_ID")
- *<span style="color:#009faf">Lat</span>*  The site latitude
- *<span style="color:#009faf">Lon</span>*  The site longitude
- *<span style="color:#009faf">startDate</span>*  The starting date for the downloaded data ("YYYY-MM-DD")
- *<span style="color:#009faf">endDate</span>*  The ending date for the downloaded data ("YYYY-MM-DD"). Default is the current date via Sys.Date()
- *<span style="color:#009faf">access_token</span>* An Earthdata access token. Requires an Earthdata profile to generate the token.
- *<span style="color:#009faf">save_dir</span>*  The save directory for the downloaded data to be placed in. For example, "C:/"

During the download some light formatting is done to extract relevant date and time information. The downloaded data can then be prepped for creating a model driver file by using the function **<span style="color:DarkOrange">nldas_prep</span>** which has the following structure:

**<span style="color:DarkOrange">nldas_prep</span>**(*<span style="color:#009faf">read_dir</span>*, *<span style="color:#009faf">write_output</span>*, *<span style="color:#009faf">save_dir</span>*)    

- *<span style="color:#009faf">read_dir</span>*  The directory containing the downloaded NLDAS data.
-  *<span style="color:#009faf">write_output</span>*  A logical value indicating whether the output should be written to disk (write_output = TRUE) or returned to the R environment (write_output = FALSE). The default value is FALSE since for most datasets this is a suitable approach; however, for very large datasets (thousands of sites) it may be easier to write files to disk instead of storing them in the workspace.
- *<span style="color:#009faf">save_dir</span>*  An optional parameter to use only when write_output = TRUE that indicates the save directory for files to be placed in. For example, "C:/".

Note, the **<span style="color:DarkOrange">nldas_prep</span>** is only creating a named list of all downloaded NLDAS data, where the names are the site ids. This is done to faciliate the later steps of creating a model driver file.

```{r echo=TRUE, results='hide', eval = FALSE}
library("StreamLightUtils")
    
#Download NLDAS data at NC_NHC
  nldas_get(
    Site_ID = "NC_NHC",
    Lat = 35.9925, 
    Lon = -79.0460,     
    startDate = "2022-01-01",
    endDate = "2024-12-31",
    access_token = access_token,
    save_dir = here::here("data", "nldas")
  )
  
#Prep the downloaded NLDAS data for creating a model driver file
  nldas_prepped <- nldas_prep(
    read_dir = here::here("data", "nldas")
  )
```

## 2. Downloading and processing NLDAS data for multiple sites

Incoming shortwave radiation data at multiple sites can be downloaded using the function **<span style="color:DarkOrange">nldas_get_bulk</span>** which has the following structure:

 **<span style="color:DarkOrange">nldas_get_bulk</span>**(*<span style="color:#009faf">site_locs</span>*, *<span style="color:#009faf">startDate</span>*, *<span style="color:#009faf">endDate</span>*, *<span style="color:#009faf">access_token</span>*, *<span style="color:#009faf">save_dir</span>*, )

- *<span style="color:#009faf">site_locs</span>*  A table with Site_ID, Lat, and Lon, and optionally startDate and endDate.
- *<span style="color:#009faf">startDate</span>*  An optional parameter. By default, if nothing is provided the function assumes that site_locs has a column that contains startDate. Alternatively, a single startDate can be provided as an argument for the download (YYYY-MM-DD).
- *<span style="color:#009faf">endDate</span>*  An optional parameter. By default, if nothing is provided the function assumes that site_locs has a column that contains endDate. Alternatively, a single endDate can be provided as an argument for the download (YYYY-MM-DD).
- *<span style="color:#009faf">access_token</span>* An Earthdata access token. Requires an Earthdata profile to generate the token.
- *<span style="color:#009faf">save_dir</span>*  The save directory for the downloaded data to be placed in. For example, "C:/".

This function can essentially be used in two ways: 1.) you can specify arguments of startDate and endDate, which may be helpful if you want to download the same timeframe of data for all sites, or 2.) specify columns of startDate and endDate in the file you pass for site_locs, which could be helpful if you want different timeframes of data for each site.

If data fails to download for a site, the **<span style="color:DarkOrange">nldas_get_bulk</span>** function will automatically check and retry downloading data for all sites with missing data. However, it is possible that data does not exist for a site. Consequently, it is prudent to confirm the successfully downloaded sites.

```{r echo=TRUE, results='hide', eval = FALSE}
#Read in a table with initial site information
  data("NC_params", package = "StreamLight") 
  sites <- NC_params

#Download NLDAS data at NC_NHC
  nldas_get_bulk(
    site_locs = sites,
    startDate = "2022-01-01",
    endDate = "2024-12-31",
    access_token = access_token,
    save_dir = here::here("data", "nldas")
  )

#Prep the downloaded NLDAS data for creating a model driver file
  nldas_prepped <- nldas_prep(
    read_dir = here::here("data", "nldas")
  )
```
